Figures to insert in text

### 2018-03-15 Chapters/sdm-architecture.tex
SDM grammar corrections

\subsection{OpenCL kernel}

There are 8 OpenCL kernels which explore differently the GPU architecture to improve performance. It is necessary because there are several GPU microarchitecture and a single kernel will never be optimal for all of them. In simplified form, OpenCL splits the tasks into workgroups which, in turn, split their part of the task into workers. The works are like threads in a computer. OpenCL specifies four levels of memory hierarchy for the GPUs: global memory, read-only memory, local memory, and private memory. The global memory and read-only memory are accessible by all workgroups, while each workgroup has its own local memory, accessible by its workers. Finally, each worker has its own private memory. The number of workers per workgroup is defined by user and must be multiple of the number of tasks.

@@

The OpenCL kernels \lstinline{single_scan3} (Listing \ref{lst:single_scan3}), \lstinline{single_scan4} (Listing \ref{lst:single_scan4}), \lstinline{single_scan5} (Listing \ref{lst:single_scan5}), \lstinline{single_scan5_unroll} (Listing \ref{lst:single_scan5_unroll}), \lstinline{single_scan6} (Listing \ref{lst:single_scan6}) explore the GPU architecture to improve the sum of the partial distances. Each workgroup calculates the distance of several bitstrings. During the distance calculation, each worker calculates the exclusive OR (XOR) between two 64-bit integers and use the built-in popcount function to count the number of ones. Then, they update an array of partial distances with their results. This array is stored in the local memory and is shared between all workers of the same workgroup. This whole step happens simultaneously in the GPU. Then, a reduction algorithm is used to sum the partial distances array in order to calculate the total distance. This reduction algorithm is also distributed between the workers and runs in $O(\log_2(\text{bs\_step}))$. Finally, the first worker of each workgroup checks whether the distance is less than or equal to the radius to include the bitstring index into the resulting array.

Some of the optimizations may not work in some GPUs, because not all their premises are valid. Before choosing a kernel, one should check whether it works property for one's specific GPU device.

@@

When two counters are added in a third counter, there may occur an overflow. It is not supposed to be a problem because, by default, each counter is a signed 32-bit integer that can store any number between -2,147,483,648 and 2,147,483,647, which means they will not overflow with less writes than $2^{31}-1$ divided by the average number of activated hard locations. For instance, when $n=1,000$, $H=1,000,000$, and $r=451$, the average number of activated hard locations is 1,000 and it would require at least one million writes before being possible to a counter to overflow.  Note also that it would be more likely to saturate the memory before any overflow.

@@

The iterated reading keeps reading until it gets exactly the same bitstring (or the number of maximum interations has been reached), i.e., it performs $\eta_{i+1} = \text{read}(\eta_i)$ and stops when $\eta_{k+1} = \eta_{k}$. If the initial bitstring is inside the critical distance of $\eta$, it will converge to $\eta$, but, if it is not, it will diverge and reach the maximum number of iterations.

The framework has both Kanerva's read and Murilo's generalized read. The generalization brings a parameter $z$, which is the exponent. In this case, the results are floating point instead of integer, which considerably reduces performance. When $z=1$, it is exactly as the Kanerva's read. When $z=0$, it is the Chada's read. We also explored how SDM would behave for different values of $z$.

There is another special read operation: the weighted reading. In the weighted reading, the value of the counters are multiplied by a weight which depends only on the distance between the reading address and the hard location address. The weight is retrieved from a lookup table of integers indexed by the distance. The rest of the read operation is exactly the same.

>>>>

\subsection{OpenCL kernels}

There are 8 OpenCL kernels which explore differently the GPU architecture to improve performance. It is necessary because there are several GPU microarchitectures and a single kernel will never be optimal for all of them. In simplified form, OpenCL splits the tasks into workgroups which, in turn, split their part of the task into workers. The works are like threads in a computer. OpenCL specifies four levels of memory hierarchy for the GPUs: global memory, read-only memory, local memory, and private memory. The global memory and read-only memory are accessible by all workgroups, while each workgroup has its own local memory, accessible by its workers. Finally, each worker has its own private memory. The number of workers per workgroup is defined by user and must be multiple of the number of tasks.

@@

The OpenCL kernels \lstinline{single_scan3} (Listing \ref{lst:single_scan3}), \lstinline{single_scan4} (Listing \ref{lst:single_scan4}), \lstinline{single_scan5} (Listing \ref{lst:single_scan5}), \lstinline{single_scan5_unroll} (Listing \ref{lst:single_scan5_unroll}), \lstinline{single_scan6} (Listing \ref{lst:single_scan6}) explore the GPU architecture to improve the sum of the partial distances. Each workgroup calculates the distance of several bitstrings. During the distance calculation, each worker calculates the exclusive OR (XOR) between two 64-bit integers and use the built-in popcount function to count the number of ones. Then, they update an array of partial distances with their results. This array is stored in the local memory and is shared between all workers of the same workgroup. This whole step happens simultaneously in the GPU. Then, a reduction algorithm is used to sum the partial distances array in order to calculate the total distance. This reduction algorithm is also distributed among the workers and runs in $O(\log_2(\text{bs\_step}))$. Finally, the first worker of each workgroup checks whether the distance is less than or equal to the radius to include the bitstring index into the resulting array.

Some of the optimizations may not work in some GPUs because not all their premises are valid. Before choosing a kernel, one should check whether it works property for one's specific GPU device.

@@

When two counters are added in a third counter, there may occur an overflow. It is not supposed to be a problem because, by default, each counter is a signed 32-bit integer that can store any number between -2,147,483,648 and 2,147,483,647, which means they will not overflow with fewer writes than $2^{31}-1$ divided by the average number of activated hard locations. For instance, when $n=1,000$, $H=1,000,000$, and $r=451$, the average number of activated hard locations is 1,000 and it would require at least one million writes before being possible to a counter to overflow.  Note also that it would be more likely to saturate the memory before any overflow.

@@

The iterated reading keeps reading until it gets exactly the same bitstring (or the number of maximum interactions has been reached), i.e., it performs $\eta_{i+1} = \text{read}(\eta_i)$ and stops when $\eta_{k+1} = \eta_{k}$. If the initial bitstring is inside the critical distance of $\eta$, it will converge to $\eta$, but, if it is not, it will diverge and reach the maximum number of iterations.

The framework has both Kanerva's read and Murilo's generalized read. The generalization brings a parameter $z$, which is the exponent. In this case, the results are floating points instead of integers, which considerably reduces performance. When $z=1$, it is exactly as the Kanerva's read. When $z=0$, it is the Chada's read. We also explored how SDM would behave for different values of $z$.

There is another special read operation: the weighted reading. In the weighted reading, the value of the counters is multiplied by a weight which depends only on the distance between the reading address and the hard location address. The weight is retrieved from a lookup table of integers indexed by the distance. The rest of the read operation is exactly the same.

### 2018-03-15 Chapters/sdm-critical-distance.tex
SDM grammar corrections

\citet[p.70]{Kanerva1988} originally predicted a \textasciitilde 500-bit distance after a point (Figure \ref{fig:kanerva-figure-7.3}). The original prediction considered that the read distance would decline when inside the critical distance and increase afterwards, converging to a \textasciitilde 500-bit distance.  At this point, each read would lead to a different, orthogonal, \textasciitilde 500-bit distance bitstring. He analyzed specifically an SDM with 1,000 bits and 10,000 random bitstrings written into it.

@@

As we ran the simulations, this one in particular struck our attention: The new distances obtained after a read operation were not perfectly predicted by the theoretical model. We have strictly followed Kanerva's configuration and, even so, we have found out some deviations from Kanerva's original theoretical analysis and the results obtained by simulation.

@@

But, when we reduced the number of random bitstrings written in the SDM from 10,000 to only 100, the results reflected very well the Kanerva's theoretical expectation (Figure \ref{fig:sdm-100w-table-7-2-10samples}). This result strengthens our hypothesis that the disparities in the computational results are due to the interaction effect of high numbers of different attractors. In Figure \ref{fig:sdm-table-7-2-steps} we can notice that, the more random bitstrings are written, the stronger the attractors.

@@

As we can easily notice, we got exactly the same curve as Kanerva. This question has been intriguing us since then and we are still looking for a more analytic explanation then merely interference from the other written attractors.

>>>>

\citet[p.70]{Kanerva1988} originally predicted a \textasciitilde 500-bit distance after a point (Figure \ref{fig:kanerva-figure-7.3}). The original prediction considered that the read distance would decline when inside the critical distance and increase afterward, converging to a \textasciitilde 500-bit distance.  At this point, each read would lead to a different, orthogonal, \textasciitilde 500-bit distance bitstring. He analyzed specifically an SDM with 1,000 bits and 10,000 random bitstrings written into it.

@@

As we ran the simulations, this one, in particular, struck our attention: The new distances obtained after a read operation were not perfectly predicted by the theoretical model. We have strictly followed Kanerva's configuration and, even so, we have found out some deviations from Kanerva's original theoretical analysis and the results obtained by simulation.

@@

But, when we reduced the number of random bitstrings written in the SDM from 10,000 to only 100, the results reflected very well the Kanerva's theoretical expectation (Figure \ref{fig:sdm-100w-table-7-2-10samples}). This result strengthens our hypothesis that the disparities in the computational results are due to the interaction effect of high numbers of different attractors. In Figure \ref{fig:sdm-table-7-2-steps} we can notice that the more random bitstrings are written, the stronger the attractors.

@@

As we can easily notice, we got exactly the same curve as Kanerva. This question has been intriguing us since then and we are still looking for a more analytic explanation than merely interference from the other written attractors.

### 2018-03-15 Chapters/sdm-performance.tex
SDM grammar corrections

Each scan on a 1,000-bits SDM with 1,000,000 hard locations executes $10^9$ bit compares through $10^9/64 = 15,625,000$ XORs and calls to the built-in popcount. So, 10,000 writes execute $10^{13}$ bit compares, while a 6-iterative reading executes $6 \cdot 10^{12}$ bit compares. The heatmap of Figure \ref{fig:cir-dist-10k-writes-kanerva} executed $3.05 \cdot 10^{15}$ bit compares. For comparison, the number of seconds since Jesus's birth is 63,639,648,000 = $6.36 \cdot 10^{10}$. The number of people who have ever lived on earth is estimated to be $1,08 \cdot 10^{11}$. There are approximtely $1.8 \cdot 10^{9}$ websites in the internet. A modern laptop can increment a counter $5 \cdot 10^{8}$ times per second. Hence, a naive implementation of SDM may take several hours --- or days --- to simply write 10,000 random bitstrings.

@@

We have created a benchmark to be able to compare the performance in different devices. So, the same performance test was executable in our devices, with results reported in tables and figures. The benchmark has 3 parts: (i) the first part consists of comparing the available OpenCL kernels to find which works best for that device; than (ii) the second part consists of comparing the linear scanner, the thread scanner, and the OpenCL scanners with the best kernel found in part one; finally, (iii) the third part consists of comparing read and write operations using the thread scanner and the OpenCL scanner with the best kernel. Each part was run for three SDM setups: (i) $n=1,000$, $r=451$, and $H=1,000,000$; (ii) $n=256$, $r=103$, and $H=1,000,000$; and (iii) $n=10,000$, $r=4850$, and $H=1,000,000$. The whole source code is available in the ``Performance test'' notebook \citep{sdmframework}. We would like to invite the reader to run this benchmark and, if possible, share the results.

Our first device was a personal MacBook Pro Retina 13-inch Late 2013 with a 2.6GHz Intel core i5 processor, 6GB DDR3 RAM, and Intel Iris GPU. It was not possible to run the 10,000-bits in this device because there was no memory available --- it needs 37.25 GB of memory. For its results, see Table \ref{tab:perf-macbook}

@@

OpenCL is a framework for writting software that executes in \emph{heterogeneous} devices \citep{munshi2009opencl}, like CPUs, GPUs, FPGA and other co-processor for hardware acceleration. Because they are heterogeneous, they may differ a lot in architecture and performance, which means there is no one-size-fits-all kernel. A \emph{kernel} is generally a small function on the code that runs in thousands of parallel threads, executing the same steps on different parts of a large vector or matrix. The slower kernel for one device may be the fastest for another device, as we will see happening in our case.

@@

The difference in which kernel is the best depends also in the SDM setup. The best kernel for the 1,000-bits SDM in the iMac 2017 was single\_scan5\_unroll with average scan time of 3.61ms; but, for the 256-bits SDM in the same device, it was single\_scan0 with average scan time of 3.03ms; while, for the 10,000-bits SDM in the same device, it was single\_scan6 with 10.96ms (see Table \ref{tab:perf-imac}).

@@

In this section, we are comparing the OpenCL scanner (with the best kernel) with the linear scanner and the threads scanner. Again, which one is faster depends on the SDM settings. In the iMac 2017, the faster scanner for a 1,000-bits SDM was the OpenCL scanner with , but for a 256-bit SDM was the threads scanner.

>>>>

Each scan on a 1,000-bits SDM with 1,000,000 hard locations executes $10^9$ bit compares through $10^9/64 = 15,625,000$ XORs and calls to the built-in popcount. So, 10,000 writes execute $10^{13}$ bit compares, while a 6-iterative reading executes $6 \cdot 10^{12}$ bit compares. The heatmap of Figure \ref{fig:cir-dist-10k-writes-kanerva} executed $3.05 \cdot 10^{15}$ bit compares. For comparison, the number of seconds since Jesus's birth is 63,639,648,000 = $6.36 \cdot 10^{10}$. The number of people who have ever lived on earth is estimated to be $1,08 \cdot 10^{11}$. There are approximately $1.8 \cdot 10^{9}$ websites on the internet. A modern laptop can increment a counter $5 \cdot 10^{8}$ times per second. Hence, a naive implementation of SDM may take several hours --- or days --- to simply write 10,000 random bitstrings.

@@

We have created a benchmark to be able to compare the performance on different devices. So, the same performance test was executable in our devices, with results reported in tables and figures. The benchmark has 3 parts: (i) the first part consists of comparing the available OpenCL kernels to find which works best for that device; than (ii) the second part consists of comparing the linear scanner, the thread scanner, and the OpenCL scanners with the best kernel found in part one; finally, (iii) the third part consists of comparing read and write operations using the thread scanner and the OpenCL scanner with the best kernel. Each part was run for three SDM setups: (i) $n=1,000$, $r=451$, and $H=1,000,000$; (ii) $n=256$, $r=103$, and $H=1,000,000$; and (iii) $n=10,000$, $r=4850$, and $H=1,000,000$. The whole source code is available in the ``Performance test'' notebook \citep{sdmframework}. We would like to invite the reader to run this benchmark and, if possible, share the results.

Our first device was a personal MacBook Pro Retina 13-inch Late 2013 with a 2.6GHz Intel core i5 processor, 6GB DDR3 RAM, and Intel Iris GPU. It was not possible to run the 10,000-bits on this device because there was no memory available --- it needs 37.25 GB of memory. For its results, see Table \ref{tab:perf-macbook}

@@

OpenCL is a framework for writing software that executes in \emph{heterogeneous} devices \citep{munshi2009opencl}, like CPUs, GPUs, FPGA and other co-processors for hardware acceleration. Because they are heterogeneous, they may differ a lot in architecture and performance, which means there is no one-size-fits-all kernel. A \emph{kernel} is generally a small function on the code that runs in thousands of parallel threads, executing the same steps on different parts of a large vector or matrix. The slower kernel for one device may be the fastest for another device, as we will see happening in our case.

@@

The difference in which kernel is the best depends also on the SDM setup. The best kernel for the 1,000-bits SDM in the iMac 2017 was single\_scan5\_unroll with average scan time of 3.61ms; but, for the 256-bits SDM in the same device, it was single\_scan0 with average scan time of 3.03ms; while, for the 10,000-bits SDM in the same device, it was single\_scan6 with 10.96ms (see Table \ref{tab:perf-imac}).

@@

In this section, we are comparing the OpenCL scanner (with the best kernel) with the linear scanner and the threads scanner. Again, which one is faster depends on the SDM settings. In the iMac 2017, the faster scanner for a 1,000-bits SDM was the OpenCL scanner with \lstinline{single_scan5_unroll}, but for a 256-bit SDM was the threads scanner.

### 2018-03-15 Chapters/sdm-validation.tex
SDM grammar corrections

In figure \ref{fig:validation-activated-hls}, we can notice that the theoretical model and the simulation matches. Hence, it seems that both the address space generation algotihm and the scan algorithm work properly. Notice that the curve is almost the same for $n=1,000$ and $n=256$. It happens because the access radius is adjust to have $p$ as close as possible to $0.001$. They are not exactly the same because their $p$ differs a little.

@@

Kanerva has calculated the intersection of two circles according to the distance between their centers. The intersection is important to understand how SDM works, because it affects directly the critical distance. When $\eta_d$ is inside the critical distance, then it will converge to $\eta$. In fact, it converges because they share a sufficient amount of hard locations, i.e., the intersection of the circle around $\eta_d$ and $\eta$ is enough to converge. For further information about the relation between the critical distance and the intersection, see \citet{brogliato2014sparse}.

We have calculated the intersection between a random bitstring (bs1) and another bitstrings (bs2) exactly $d$ bits away. The former (bs1) is just a random bitstring. The latter (bs2) was generated randomly flipping $d$ bits of bs1. The code is available in the ``Kanerva's Figure 1.2'' notebook \citep{sdmframework}.

@@

It worked as if we had splitted SDM into $k$ regions with low intersection between two of them. So, as the interference is minimal, they work like independent SDMs. The major disadvantage of this approach is that memory capacity may be reached faster.

>>>>

In figure \ref{fig:validation-activated-hls}, we can notice that the theoretical model and the simulation matches. Hence, it seems that both the address space generation algorithm and the scan algorithm work properly. Notice that the curve is almost the same for $n=1,000$ and $n=256$. It happens because the access radius is adjusted to get $p$ as close as possible to $0.001$. They are not exactly the same because their $p$ differs a little.

@@

Kanerva has calculated the intersection of two circles according to the distance between their centers. The intersection is important to understand how SDM works because it affects directly the critical distance. When $\eta_d$ is inside the critical distance, then it will converge to $\eta$. In fact, it converges because they share a sufficient amount of hard locations, i.e., the intersection of the circle around $\eta_d$ and $\eta$ is enough to converge. For further information about the relation between the critical distance and the intersection, see \citet{brogliato2014sparse}.

We have calculated the intersection between a random bitstring (bs1) and another bitstring (bs2) exactly $d$ bits away. The former (bs1) is just a random bitstring. The latter (bs2) was generated randomly flipping $d$ bits of bs1. The code is available in the ``Kanerva's Figure 1.2'' notebook \citep{sdmframework}.

@@

It worked as if we had split SDM into $k$ regions with low intersection between two of them. So, as the interference is minimal, they work like independent SDMs. The major disadvantage of this approach is that memory capacity may be reached faster.
